Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-18 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

![Schematic of k-fold cross validation. The dataset is randomly divided into k stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all k folds.](images/k-fold_fig.png){#fig-k-fold width="6000"}

@@

Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity. These findings emphasize the critical role of climate-hydrology interactions in shaping groundwater contributions to streamflow. Inverse trends between temperature and BFI suggest that further warming could reduce groundwater contributions to streamflow. Coincident trends in precipitation and BFI further underscore the importance understanding recharge processes in arid and semi-arid landscapes.

While there has been an increase of stream gages in Arizona to comply with regulatory imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses. New, non-USGS stream gages are typically installed by flood control districts for flood flow monitoring, and do not accurately track base flow. New USGS stream gages have been added over the past 10 years to address instream flow rights and to improve the density of monitoring in the future.

>>>>

![Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold width="6000"}

@@

Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity. These findings emphasize the critical role of climate-hydrology interactions in shaping groundwater contributions to streamflow. Inverse trends between temperature and BFI suggest that further warming could reduce groundwater contributions to streamflow. Coincident trends in precipitation and BFI further underscore the importance of understanding recharge processes, especially in arid and semi-arid landscapes, where precipitation events play key roles in replenishing groundwater.

However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus.There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act. Even so, many of these gages are not suited for base-flow studies because they emphasize flood monitoring and lack the ability to accurately measure low-flow dynamics. New, non-USGS streamgages are typically installed by flood control districts (e.g. the ALERT system) and are designed for tracking flood flows rather than base flow. New USGS streamgages have been added over the past decade to address instream flow rights and to improve the density of monitoring in the future.
